MSD panels 1, 2 and 3 Explant Sup

### 2016-08-19 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Updated code to include panel 1 data, added in a bunch of sanity checks.

Caveats:
Due to limited sample volume, we had to make the following omissions:

@@

Due to a possible pipetting error, we were not able to use the data from the standards that we ran for panel 1. Instead, we used the standards from a previous run of the same panel to calculate concentrations for the experimental samples(see explant_sup_README for details).

@@

#remember that this data table contains data from the panel 1-Test from 28Mar16 (Plate_2BF6AAN447). I am only going to use the Standards data from that plate and not the standards from either of the plates for panel 1 run 10Aug16.

# HOWEVER: When we did panel 1-TEST, the kit that we used was missing the Ab for MIP1B, so we didn't include that one the plate, so we don't have any standard data for that analyte from this plate. So, we will need to use the standard data from the 10Aug16 run of panel 1 for that analyte, or, not use it.

#annoyingly, for Plate_2BF6AAN447, the standards are not named "STD-01" etc in the sample names column but rather "S001". I'm going to change them to match the other panels.

panel_1 <- read.csv("../raw_data/10Aug16_explant_sup_panel_1_plate_1_and_2.csv", skip = 1)

TestStdsToUse <- panel_1 %>%
  filter(Sample.Group == "Standards")%>%
  filter(Plate.Name == "panel_1_TEST")

#fixing the sample names!
TestStdsToUse$Sample<-str_replace(TestStdsToUse$Sample, "S0","STD-")

#filtering out just the standards for MIP1B from the 11Aug16 panel 1 plates (don't have data for that analyte in the test panel). The greek letters got all messed up since they went thru excel, but I know which is which based on the Spot (in the well) that they are associated with.

StdsToUse<- panel_1 %>%
  filter(Sample.Group == "Standards", Spot == "4")

#now filter out all the standards from the panel 1 data and add back in just the standards I want (TestStdsToUse and StdsToUse)

panel_1NoStds<-panel_1 %>%
  filter(Sample.Group != "Standards")

panel_1_Stds_corrected <- rbind(panel_1NoStds,StdsToUse,TestStdsToUse )

# Data from panels 2 and 3
panel_2<-read.csv("../raw_data/10Aug16_explant_sup_panel_2_plate_1_and_2.csv", skip = 1)

panel_3<-read.csv("../raw_data/11Aug16_explant_sup_panel_3_plate_1_and_2.csv", skip = 1)

#Combining the data from all 3 panels

allPanels<- rbind(panel_1_Stds_corrected, panel_2,panel_3)

@@

```{r fixing panel 1 standards, echo=FALSE, message = FALSE, warning = FALSE}

#just the standards from all the panels

##SANITY CHECK
# panel 1: 8std x 2reps x 1analytes x 2 plates = 32
# panel 1 TEST: 8std x 2reps x 8analyte x 1plate = 128
# panel 2: 8std x 2reps x 8analytes x 2 plates = 256
# panel 3: 8std x 2reps x 3analytes x 2 plates = 96

# 32+128+256+96=512

allPanelsStds<-allPanels %>%
  filter(Sample.Group == "Standards")%>%
  group_by(Plate.Name)
```

@@

The R^2 values for goodness-of-fit for the standard curves were all >0.99
```{r panel 2 and 3 standards, echo=FALSE, message = FALSE, warning = FALSE}

#I want to ignore panel 1 data right now since we might get a better set of standards to use with it.

panel_2_and_3<- rbind(panel_3, panel_2)

panel_2_and_3Stds<-filter(panel_2_and_3, Sample.Group == "Standards")

#which standards are out of det. range?
standardsOutOfRange<-panel_2_and_3Stds %>%
  filter(Detection.Range != "In Detection Range")%>%
  arrange(Plate.Name,Sample, Assay, Detection.Range)%>%
  select(Plate.Name,Sample, Assay, Detection.Range)

stdFit<- panel_2_and_3Stds %>% 
  arrange(Sample,Assay)%>%
  filter(Detection.Range != "Below Fit Curve Range")%>%
  select(Sample, Assay,Detection.Range, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)

ggplot(stdFit, aes(x = Sample, y = X..Recovery))+
  geom_point(aes(color = Sample, shape = Detection.Range ),size = 2, alpha = 0.5)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "In Detection Range" = 16, "Below Detection Range" = 0))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(0, 175))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = rel(0.8)))

#checking the R squared goodness-of-fit for the standard curves
rSquared<-stdFit%>%
  group_by(Assay)%>%
  summarize(R_Squared = mean(Fit.Statistic..RSquared))

@@

```{r clean data, echo=FALSE, message = FALSE, warning = FALSE}

@@

panel_2_and_3_noStds <- panel_2_and_3 %>%
  filter(Sample.Group!="Standards")

# I want to be able to group the data by timepoint and donor and virus so I need to separate the sample name and make new columns identifying them that way

donorTimeVirus<-colsplit(panel_2_and_3_noStds$Sample,"-", c("Donor","Timepoint","Virus") )

#bind the split col back to panel_2_and_3

panel_2_and_3_noStds<-cbind(donorTimeVirus, panel_2_and_3_noStds)

#Make things factors so they plot nicely
panel_2_and_3_noStds$Donor <- as.factor(panel_2_and_3_noStds$Donor)

panel_2_and_3_noStds$Calc..Concentration <-as.numeric(panel_2_and_3_noStds$Calc..Concentration)

panel_2_and_3_noStds$Virus <- as.factor(panel_2_and_3_noStds$Virus)

panel_2_and_3_noStds$Virus<-factor(panel_2_and_3_noStds$Virus, levels = c("M", "V1","V2"))

panel_2_and_3_noStds$Timepoint<-as.factor(panel_2_and_3_noStds$Timepoint)
```

@@

#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep

replicateSummary <- panel_2_and_3_noStds %>%
group_by(Plate.Name,Sample, Assay)%>%
summarize(reps = n())

@@

@@

#Which were below detection range?
panel_2_and_3_noStds_Below <- panel_2_and_3_noStds %>%
  filter(Detection.Range == "Below Fit Curve Range" & Detection.Range != "Below Detection Range")

panel_2_and_3_noStds_noBelow <- panel_2_and_3_noStds %>%
  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")

#averaging the replicates
avgConcpanel_2_and_3_noStds_noBelow <- panel_2_and_3_noStds_noBelow %>%
group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
summarize(avgConc = mean(Calc..Concentration))%>%
  ungroup()

@@

numberInSets<-avgConcpanel_2_and_3_noStds_noBelow %>%
         group_by(Donor,Timepoint,Assay)%>%
  select(Sample,Donor, Timepoint,Assay)%>%
  summarize(numInSet = n())

@@

avgConcpanel_2_and_3_noStds_noBelow <- merge(avgConcpanel_2_and_3_noStds_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))

@@

avgConcpanel_2_and_3_noStds_noBelow_fullSet <- avgConcpanel_2_and_3_noStds_noBelow %>%
  filter(numInSet != 1)

@@

@@

Also note:
Some of the samples had one rep that was below detection and another that was above. For these samples I used the remaining rep as the "average" for that sample.

@@

#make a list of the analytes
analyteList<-as.list(levels(avgConcpanel_2_and_3_noStds_noBelow$Assay))

@@

plotAnalytes <- function (Analyte){
   avgConcpanel_2_and_3_noStds_noBelow_fullSet %>%
     filter(Assay == Analyte)%>%

     ggplot( aes(x = Virus, y = avgConc))+

   geom_point(aes(color = Donor ),size = 1.5)+
  geom_line(aes(group = Donor, color = Donor)) +

@@

  facet_wrap(~Timepoint, scales = "free_x")+

@@

lapply(analyteList, FUN = plotAnalytes)

pander(panel_2_and_3_noStds %>%
  select(Sample, Assay, Calc..Concentration, Detection.Range))

>>>>

## Caveats

Due to limited sample volume, we had to make the following omissions:

@@

Due to a possible pipetting error, we were not able to use the data from the standards that we ran for panel 1. Instead, we used the standards from a different run of the same panel to calculate concentrations for the experimental samples(see explant_sup_README for details).

@@

#remember that this data table contains data from the 17Aug16_HVE_sup_panel_1_plate_1. I am only going to use the Standards data from that plate and not the standards from either of the plates for panel 1 run 10Aug16 since we had what appears to be a pipetting error (2nd replicate is more concentration for all wells).

explant_panel_1_and_HVE_panel_1_plate_1 <- read.csv("../raw_data/10Aug16_explant_sup_panel_1_plate_1_and_2.csv", skip = 1)

#Filter to ONLY keep the standards from the plate called "HVE_sup_panel_1_plate_1"

StdsToUse_explant_panel_1 <- explant_panel_1_and_HVE_panel_1_plate_1 %>%
  filter(Sample.Group == "Standards")%>%
  filter(Plate.Name == "HVE_sup_panel_1_plate_1")

#Since I told Discovery Workbench software that I wanted to use the standards from the 17Aug16_HVE_sup_panel_1_plate_1 as "Global Standards" (i.e. use them to calculate concentrations for all plates in the experiment)it seems to have provided me with 3 copies of those samples in the data table (1 for each plate I guess). I only need data from the unique entries, so I will filter for just unique rows in this df to exclude copies.

# should have 8 stds x 9 analytes x 2 reps = 144 obs
StdsToUse_explant_panel_1<-unique(StdsToUse_explant_panel_1)

#Sanity check: there should only be 2 entries for each unique instance of Sample (STD-0x) and assay. 
checkStdReps<-StdsToUse_explant_panel_1 %>%
         group_by(Sample,Assay)%>%
         summarise(n())

#Remove the 17Aug16_HVE_sup_panel_1_plate_1 non-STD data from the df and leave ONLY explant panel 1 data. Remaining should be:
#94 wells from plate 1 + 30 wells from plate 2 x 9 analytes = 1116 observations

explant_panel_1 <-explant_panel_1_and_HVE_panel_1_plate_1 %>%
  filter(Plate.Name != "HVE_sup_panel_1_plate_1")

#now remove the unwanted Std samples from the explant data. There should be 32 Std wells x 9 analytes = 288
# 1116 total - 288 = 828 observations

No_Stds_explant_panel_1 <-explant_panel_1 %>%
  filter(Sample.Group != "Standards")

#now add in the unique Stds from 17Aug16_HVE_sup_panel_1_plate_1

corrected_Stds_explant_panel_1 <- rbind(No_Stds_explant_panel_1,StdsToUse_explant_panel_1)

# Data from panels 2 and 3

#panel 2 should have 123 wells (left out one rep) x 8 analytes = 984
explant_panel_2<-read.csv("../raw_data/10Aug16_explant_sup_panel_2_plate_1_and_2.csv", skip = 1)

#panel 3: 120 wells (left out 2 samples, 4 reps) x 8 analytes =360

explant_panel_3<-read.csv("../raw_data/11Aug16_explant_sup_panel_3_plate_1_and_2.csv", skip = 1)

#Combining the data from all 3 panels: 972 +984 + 360= 2316

explant_all_panels<- rbind(corrected_Stds_explant_panel_1, explant_panel_2,explant_panel_3)

@@

@@

The R^2^ values for goodness-of-fit for the standard curves were all >0.99

```{r panel 2 and 3 standards, echo=FALSE, message = FALSE, warning = FALSE}

explant_stds_all_panels<-filter(explant_all_panels, Sample.Group == "Standards")

#which standards are out of det. range?
standardsOutOfRange<-explant_stds_all_panels %>%
  filter(Detection.Range != "In Detection Range")%>%
  select(Plate.Name,Sample, Assay, Detection.Range, X..Recovery, Calc..Concentration)

#looking at which samples have STDs below the fit curve or below the detection range (i.e. you can't get numbers for % recovery) 
explant_stds_all_panels_below<- explant_stds_all_panels%>%
  filter(Detection.Range == "Below Fit Curve Range" | Detection.Range == "Below Detection Range") %>%
  group_by(Sample)%>%
  summarise("Number of Samples Below Fit Curve Range/Detection Range" = n())

#The samples that actually give me #'s for percent recovery ( i.e. all but the "below fit curve range" and "below detection range" ones)

explant_stds_all_panels_percent_recov<- explant_stds_all_panels %>% 
  arrange(Sample,Assay)%>%
  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range!= "Below Detection Range")%>%
  select(Plate.Name,Sample, Assay,Detection.Range, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)

#plotting percent recovery

ggplot(explant_stds_all_panels_percent_recov,aes( x = Sample, y = X..Recovery)) +
  geom_point(aes(shape = Detection.Range ),
             size = 2, alpha = 0.5)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, 
                                "In Detection Range" = 16, 
                                "Below Detection Range" = 0))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(0, 250))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = rel(0.8)))

#checking the R squared goodness-of-fit for the standard curves
rSquared<-explant_stds_all_panels_percent_recov%>%
  group_by(Assay)%>%
  summarize(R_Squared = mean(Fit.Statistic..RSquared))

@@

## Caveats about Standards

* Standards with values below the detection range (<2.5x the standard deviation of the lowest standard) are give unreliable concentrations, so I did not include those in the plots.

* When standards are below the curve fit (outside the range of the standard curve), the software does not calculate concentrations so you can't get a percent recovery.

* Most of the Standards that fell into either of these categories were from STD-07 or STD-08 except for two that were STD-06:

`r pander(explant_stds_all_panels_below)r`

```{r break up sample name, echo=FALSE, message = FALSE, warning = FALSE}

@@

explant_all_panels_noStds <- explant_all_panels %>%
  filter(Sample.Group!="Standards")

# I want to be able to group the data by timepoint and donor and virus so I need to break up components of the the sample name and make new columns with just the Donor, Timepoint and Virus

#Split samples names on the "-"
donorTimeVirus<-colsplit(explant_all_panels_noStds$Sample,"-", c("Donor","Timepoint","Virus") )

#bind the split col back to allPanels

explant_all_panels_noStds<-cbind(donorTimeVirus, explant_all_panels_noStds)

#Make things factors so they plot nicely
explant_all_panels_noStds$Donor <- as.factor(explant_all_panels_noStds$Donor)

explant_all_panels_noStds$Calc..Concentration <-as.numeric(explant_all_panels_noStds$Calc..Concentration)

explant_all_panels_noStds$Virus <- as.factor(explant_all_panels_noStds$Virus)

explant_all_panels_noStds$Virus<-factor(explant_all_panels_noStds$Virus, levels = c("M", "V1","V2"))

explant_all_panels_noStds$Timepoint<-as.factor(explant_all_panels_noStds$Timepoint)
```

@@

#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep (sample 324-T2-V1 where I ran out of sample before I could do the 2nd rep)

replicateSummary <- explant_all_panels_noStds %>%
group_by(Plate.Name,Sample, Assay)%>%
summarize(reps = n())

@@

## Explant Supernatents: Detection Range

```{r checking detection range, echo=FALSE, message = FALSE, warning = FALSE}

#What number of samples are in each category of detection?
explant_all_panels_noStds_Below <- explant_all_panels_noStds%>%
 group_by(Assay, Detection.Range)%>%
  summarise(Number_of_Samples = n())

#Let's plot that

ggplot(explant_all_panels_noStds_Below, aes(x = Assay, y=Number_of_Samples))+
         geom_point(aes(color = Detection.Range),size = 2)+
  theme(axis.text.x = element_text(angle=25))+
  ggtitle("Number of Samples in each Detection Range per Analyte")

@@

explant_all_panels_noStds_noBelow <- explant_all_panels_noStds %>%
  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")

#averaging the replicates
avgConc_explant_all_panels_noStds_noBelow <- explant_all_panels_noStds_noBelow %>%
group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
summarize(avgConc = mean(Calc..Concentration))%>%
  ungroup()

@@

numberInSets<-avgConc_explant_all_panels_noStds_noBelow %>%
         group_by(Donor,Timepoint,Assay)%>%
  select(Sample,Donor, Timepoint,Assay)%>%
  summarize(numInSet = n())

@@

avgConc_explant_all_panels_noStds_noBelow <- merge(avgConc_explant_all_panels_noStds_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))

@@

avgConc_explant_all_panels_noStds_noBelow_fullSet <- avgConc_explant_all_panels_noStds_noBelow %>%
  filter(numInSet != 1)

@@

Black points show the means.

@@

Also note:
Some of the samples had one rep that was below detection and another that was above. For these samples I used the remaining rep as the "average" for that sample.
```{r correcting analyte names, echo=FALSE, message = FALSE, warning = FALSE}

#excel messes up the greek letters of the analytes so I'm fixing them here.

#unicode greek lower case alpha = U03B1

@@

#make a list of the analytes
analyteList<-as.list(levels(avgConc_explant_all_panels_noStds_noBelow$Assay))

@@

plotAnalytes <- function (Analyte){
   avgConc_explant_all_panels_noStds_noBelow_fullSet %>%
     filter(Assay == Analyte)%>%

     ggplot( aes(x = Timepoint, y = avgConc))+

   geom_point(aes(color = Donor ),size = 1.5, alpha = 0.5)+
  geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
    stat_summary(fun.y = mean, geom = "point")+
    stat_summary(fun.y = mean, geom = "line",aes(group = 1))+

@@

  facet_wrap(~Virus, scales = "free_x")+

@@

lapply(analyteList, FUN = plotAnalytes)
```
